---
title: "VCF Trio Structural Variant Analysis"
output:
  html_document:
    theme: readable
    highlight: zenburn
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


<p style="text-align: center;">Structural variant screening and functional annotation of a parent–offspring trio</p>
## BASH

### a. Summarize the various combination of variants, their location, type and function.

1. Download the data file for the bioinf-m2 group project.

```{bash}
# get-data bioinf-m2_group-project 
```

2. Begin typing this path and press Tab to view the files or folders available in that directory.

```{bash}
# cp /data/bioinf-m2_group-project/
```
3. Show the line(s) in the VCF file that include the word CHROM. This line typically corresponds to the header displaying the column names of the VCF.
In this project, the DellyVariation.vcf file was used.

```{bash}
grep CHROM DellyVariation.vcf 
```
4. Deleting the columns from other projects and creating a new file that includes only the required data and columns (i.e., removing HG00731, HG00732, HG00733, NA19238, NA19239, NA19240, and keeping HG00512, HG00513, HG00514).

```{bash}
awk -v OFS="\t" '!/##/ {$13=$14=$15=$16=$17=$18="";print}' DellyVariation.vcf > DellyTrio1.vcf|sed 's/^\s\+//g'
```

5. The columns were removed.

```{bash}
grep CHROM DellyTrio1.vcf 
```
6. Number of variants present in the VCF file.

```{bash}
wc -l DellyTrio1.vcf 
```

<div style="background-color:#f0ffff; padding:10px; border-left:5px solid #1E90FF; border-radius:5px; margin-bottom:10px;">
The dataset contains 17312 genetic variations identified across the analyzed samples.
</div>


7. Obtain the exact number of variants in the VCF file, excluding the header lines.

```{bash}
awk '$0!~/^#/' DellyTrio1.vcf | wc -l 
```
8. FILTER column and determine the number of unique entries, showing how many variants are PASS compared to lowQual.

```{bash}

awk '$0!~/^#/ {print $7}' DellyTrio1.vcf | sort | uniq -c | sort -n
```

<div style="background-color:#f0ffff; padding:10px; border-left:5px solid #1E90FF; border-radius:5px; margin-bottom:10px;">
Out of all variants, 17152 passed quality filters (PASS) and 159 are low quality (LowQual).
</div>


9. Select the variants that are marked as “PASS” in the FILTER column.

```{bash}
awk '$0!~/^#/ && $7=="PASS"' DellyTrio1.vcf | head -2
```
10. Select multi-allelic variants. 
```{bash}
awk '$0!~/^#/ && $5~/,/' DellyTrio1.vcf
```

<div style="background-color:#f0ffff; padding:10px; border-left:5px solid #1E90FF; border-radius:5px; margin-bottom:10px;">
No multi-allelic variants were detected in the VCF file.
</div>

11. Select bi-allelic SNPs.
```{bash}
awk '$0!~/^#/ && $5!~/,/ && length($4)==1 && length($5)==1' DellyTrio1.vcf

awk '$0!~/^#/ && $5!~/,/ && length($4)==1 && length($5)==1' DellyTrio1.vcf | sort | uniq -c
```

<div style="background-color:#f0ffff; padding:10px; border-left:5px solid #1E90FF; border-radius:5px; margin-bottom:10px;">
No simple bi-allelic SNPs were found in the VCF file.
</div>


12. Select bi-allelic variants. 
```{bash}

awk '$0!~/^#/ && $5!~/,/' DellyTrio1.vcf | head -2

awk '$0!~/^#/ && $5!~/,/' DellyTrio1.vcf | cut -f4,5 | head -2

```
<div style="background-color:#f0ffff; padding:10px; border-left:5px solid #1E90FF; border-radius:5px; margin-bottom:10px;">
The VCF file contains variants, including deletions (DEL) and insertions (INS). No simple SNPs were detected.
</div>

### b. Identify de novo mutation in the trio.
13. Deletes all lines that contain ‘N’
```{bash}
awk 'BEGIN{OFS="\t"} /^#/ {print; next} $4 != "N"' DellyTrio1.vcf > CleanDellyTrio1Pass.vcf 
head -2 CleanDellyTrio1Pass.vcf
```

14.  Identifying de novo mutations.
```{bash}
awk '$0!~/^#/ && $11 ~ /^0\/0/ && $10 ~ /^0\/0/ && $12 !~ /^0\/0/ && $12 !~ /^\.\/\./' CleanDellyTrio1Pass.vcf > family.denovo.vcf 


```

<div style="background-color:#f0ffff; padding:10px; border-left:5px solid #1E90FF; border-radius:5px; margin-bottom:10px;">
When we first looked for de novo mutations, three sites were found, but all contained ambiguous bases (N). After removing these ambiguous positions, *no de novo* mutations were detected.
</div>

15. Start by creating a table of genotypes for variants marked as "PASS".
```{bash}
awk '$0!~/^#/ && $7=="PASS" {s=$1"\t"$2 ; for (i = 10; i <= NF; i++) {split($i,x,":") ; s=s"\t"x[1]}; print s}' CleanDellyTrio1Pass.vcf > genotypes.txt
```

16. Determining which sample corresponds to the child, father and mother.

16.1. **Child**
```{bash}
for nn in 3,4 3,5 4,5 ; do echo "----------> $nn" ; cut -f $nn genotypes.txt | awk '($1=="0/0" && $2=="1/1") || ($2=="0/0" && $1=="1/1")' | wc -l ; done
```

<div style="background-color:#f0ffff; padding:10px; border-left:5px solid #1E90FF; border-radius:5px; margin-bottom:10px;">
The child corresponds to the sample whose genotypes would differ from both parents. In our dataset, after checking the columns, HG00514 was identified as the child.
</div>

16.2. **Mother and father.** Finding the different information in the FORMAT column
 
```{bash}
grep FORMAT DellyVariation.vcf 

```

16.3. Creating a table of reads on the chromosomes and removing entries containing 0 0 0 [Adapting genotype table code]

```{bash}
awk '$0!~/^#/ && $7=="PASS" {s=$1"\t"$2 ; for (i = 10; i <= NF; i++) {split($i,x,":") ; s=s"\t"x[6]}; print s}' CleanDellyTrio1Pass.vcf > ReadCounts.txt

awk '!($3==0 && $4==0 && $5==0)' ReadCounts.txt > CleanReadCounts.txt

```
16.4.  Comparing read counts between samples

```{r}
library(stringr)
readcounts <- read.delim('CleanReadCounts.txt',header = F)
colnames(readcounts) <- c("Chr","Position","HG00512","HG00513","Child")
chromX <- readcounts[readcounts$Chr == "chrX", ]
head(chromX)

sum(chromX$HG00512)
sum(chromX$HG00513)
sum(chromX$Child)
```

<div style="background-color:#f0ffff; padding:10px; border-left:5px solid #1E90FF; border-radius:5px; margin-bottom:10px;">
By looking at the read counts of the samples, we can identify that HG00512 must be the father due to having a lower read count on the X chromosome compared to the other two samples. This is because males have 1 X chromosome while females have 2 X chromosomes. We can also conclude that HG00513 and the child must be female.
</div>

16.5. Making bar chart of read counts on chromosome X for visualisation

```{r}
library(dplyr)
library(RColorBrewer)
library(viridis)
library(ggplot2)

readcounts <- c(sum((chromX$HG00512) / 251), sum((chromX$HG00513) / 251), sum((chromX$Child) / 251))
samples <- c("HG00512", "HG00513", "Child")
readcountdf <- data.frame(Sample = factor(samples, levels = c("HG00512", "HG00513", "Child")), Reads = readcounts, Sex = c("Male","Female","Child"))

barplot(readcountdf$Reads, names.arg = readcountdf$Sample, main = 'X Chromosome Reads', xlab = 'Samples', ylab = 'Reads', ylim = c(0, 4500), col =brewer.pal(n = 1, name = "RdBu"), las = 1, cex.names = 0.7)

```


## ANNOVAR 

### c. Annotate the variants of interest 

17. The VCF file was converted into Avinput format for analysis with ANNOVAR.
```{bash}
# perl convert2annovar.pl -format vcf4 CleanDellyTrio1Pass.vcf > CleanedDT1Pass.avinput
```

18. The ANNOVAR tool and its databases (hg38 human) required for analysis were obtained from the official [ANNOVAR website](https://annovar.openbioinformatics.org/en/latest/user-guide/download/).

```{bash}

#perl annotate_variation.pl -buildver hg38 -downdb -webfrom annovar refGeneWithVer humandb/
#perl annotate_variation.pl -buildver hg38 -downdb cytoBand humandb/
#perl annotate_variation.pl -buildver hg38 -downdb -webfrom annovar gnomad41_genome humandb/
#perl annotate_variation.pl -buildver hg38 -downdb -webfrom annovar clinvar_20250721 humandb/
```

<div style="background-color:#f0ffff; padding:10px; border-left:5px solid #1E90FF; border-radius:5px; margin-bottom:10px;">
These databases allow us to obtain important scores for the analysis of the variants of interest. According to ANNOVAR, each corresponds to: `refGeneWithVer`: Gene annotation, `cytoBand`: To identify Giemsa-stained chromosomes bands, `gnomad41_genome`: Genome Aggregation Database version 4.1 whole-genome data, `clinvar_20250721`: CLINVAR database with Variant Clinical Significance.
</div>


19. Run ANNOVAR to get a csv file with all the variants.

```{bash}

# perl table_annovar.pl CleanedDT1Pass.avinput
# humandb/ 
# -buildver hg38 -out hg38test -remove -protocol 
# refGeneWithVer,cytoBand,gnomad41_genome,clinvar_20250721 
# -operation gx,r,f,f 
# -nastring . -csvout -polish -xreffile example/gene_fullxref.txt
```
## R 

20. Running the csv file in R

```{r}
library(tidyverse)

df <- read.csv(file="annovar_output.csv", header = T)

filter_variants <- c("nonframeshift insertion", "frameshift deletion","nonframeshift deletion")

data_filter <-filter(df, ExonicFunc.refGeneWithVer %in% filter_variants) %>% select(Chr, Start, End)

head(data_filter,8)
```

<div style="background-color:#f0ffff; padding:10px; border-left:5px solid #1E90FF; border-radius:5px; margin-bottom:10px;">
The results were filtered taking into account the type of mutation present, which yielded a total of eight positions across chromosomes 1, 2, 3, 4, 9, 14, and 18.
</div>


21. Searching for combinations of variants at the eight identified positions in the avinput file.


```{r}
df_avinput <- read.table(file="CleanedDT1Pass.avinput", header = F) %>% select(V1,V2,V3,V4,V5,V6)
colnames (df_avinput) = c ("Chr","Start","End","Ref","Alt","Genotype")

select_chrom <- c("177960892", "108259009", "214780779", "195791241", "1093539", "76703460", "20060290", "2906779")
data_chrom <-filter(df_avinput, Start %in% select_chrom) 

head(data_chrom,8)
```

<div style="background-color:#f0ffff; padding:10px; border-left:5px solid #1E90FF; border-radius:5px; margin-bottom:10px;">
The result shows that six genes are heterozygous and two are homozygous.
</div>

22. Bar chart of chromosome location

```{r}

ChromosomeOrder <- c("chr1","chr2","chr3","chr4","chr5","chr6","chr7","chr8","chr9","chr10","chr11","chr12","chr13","chr14","chr15","chr16","chr17","chr18","chr19","chr20","chr21","chr22","chrX")
Chrom <- table(df$Chr)
Chrom_ordered <- Chrom[ChromosomeOrder]

barplot(Chrom_ordered,
        main = 'Variant Locations',
        xlab = 'Chromosome',
        ylab = 'Frequency',
                ylim = c(0, 500), 
        col =brewer.pal(n = 8, name = "RdBu"),
        las = 2,           
        cex.names = 0.7)

chrm_perc <- table(df$Chr)
chrm_percent <- round(chrm_perc / sum(chrm_perc) * 100, 1)

```

23. Pie chart for types of variants

```{r}
Variants <- table(df$ExonicFunc.refGeneWithVer)
filter_var <- data.frame(Variants[-c(1,5)])
filter_var$Percent <- round(filter_var$Freq / sum(filter_var$Freq) * 100, 1)
head(filter_var)
ggplot(filter_var, aes(x="", y=Freq, fill=Var1)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void() + 
  scale_fill_brewer(palette="Accent") 


```


24. Frequency table for region function.
```{r}
regions <- table(df$Func.refGeneWithVer)
regions <- sort(regions, decreasing = T)
regionsdf <- data.frame(Region = names(regions), Freq = as.numeric(regions))

regionsdf$Percent <- (regionsdf$Freq / sum(regionsdf$Freq)) * 100

print(regionsdf)

```


25. Finding out how many insertions and deletions are in the VCF file

```{bash}

egrep -c "SVTYPE=INS" CleanDellyTrio1Pass.vcf
egrep -c "SVTYPE=DEL" CleanDellyTrio1Pass.vcf

```

<div style="background-color:#f0ffff; padding:10px; border-left:5px solid #1E90FF; border-radius:5px; margin-bottom:10px;">
Exonic variants can directly affect protein coding regions, whereas intronic variants may influence regulatory sequences such as enhancers or splice sites.
</div>
26. Pie chart for insertions and deletions

```{r}
sections <- c(9444, 5315)
labels <- c("Deletions","Insertions")
percentages <- round(sections/sum(sections)*100, digits=2)
labels <- paste(labels, percentages)
labels <- paste(labels, "%", sep="")
pie(sections, labels=labels, col=rainbow(length(labels)), main="Variant Types")
```


